Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-04 paper/Intuition.tex
some modifications of backgrounda

Given these invariants, we codify our intuition by showing that
Theorem~\textit{storeAtomicity} follows naturally from these
invariants.  A store may happen only when leaf cache is in \Mo{} state
for that address (Invariant~\textit{processReq}). Leaf caches cannot
be ancestors of any other node, so by Invariant
\textit{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
other leaf caches will be in \In{} state preventing the possibility of
load and store requests for the same address (by Invariant
\textit{processReq}), showing Definition~\textit{uniqStTimes} holds.

Given Invariant~\textit{noReqAgain}, once a response has been produced
for a request $i$, the cache may subsequently only process $j^{th}$
request such that $j > i$. Thus there is at most one response to each
request, satisfying Definition~\textit{uniqRespLabels}. A load request
may be processed only if a cache is at least in \Sh{} state (Invariant
\textit{processReq}). As leaf caches have no children, a leaf cache in
at least in \Sh{} state for address $a$ is \clean for that address. A
load response supplies the data for that address present in the cache,
which, by Invariant~\textit{lastestValue}, contains the ``latest
value''. Thus, the final condition for Theorem~\textit{storeAtomicity}
is also satisfied. All that remains is to show that these properties hold.

@@

We can directly show that Invariants~\textit{processReq} and
\textit{noReqAgain} hold at the initial configuration and are
preserved through each atomic transition of our protocol
(Figure~\ref{trans}). Invariants~\textit{uniqM} and
\textit{latestValue} correctness rely on properties of the node
hierarchy. We can prove both using the following invariants on
parent-child relationships.

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \fullref{conservative} and
\textit{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \fullref{conservative} and the first part
\textit{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

@@

To prove Invariant \fullref{conservative} at some time $t$, consider the case
when a child cache $c$ has sent to or received from the parent $p$ some message
before $t$. Consider the last such message $m$ sent by or received at $c$ at
$t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
from $c$ after sending $m$ in order to not violate Invariant \fullref{noCross}.
Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
sending $m$ in order to not violate Invariants \fullref{noCross} and
\textit{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
the response messages sync up the next-state according to the transitions in
Figure \ref{trans}). Since $p$ can not receive any message between $t''+1$ and
$t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
between $t'+1$ and $t$, Invariant \fullref{conservative} holds at $t$. If no
message has been sent or received by $c$ before $t$, then $p$
wouldn't have received any message sent by $c$ before $t$ as it means that the
message must be sent by $c$ before $t$. Since the initial states of the
directory and the child's state are both set to \In{}, the same argument holds
(in which the parent can only upgrade), thus maintaining Invariant
\textit{conservative}.

@@

Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
other response can be received by the parent at time $t_p$ such that $t+1 \le
t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
directory for $c$. Similarly, no response can be sent by the child at time
$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
child upgrades its state at $t'$. Using arguments similar to the proof of
Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
state for address $a$ when a response is in transit from $p$ to $c$.
Similarly, we can prove that no cache can be in \Mo{} state for address $a$
when a response is in transit for address $a$ from $c$ to $p$. Since data is in
transit at the same time as when the response is in transit, Invariant
\textit{noTransitWrite} follows.

@@

%Invariant~\textit{conservative} can be proved by

@@

%Invariant~\textit{conservative} at any time $t$ for child cache $c$ and parent

>>>>

Given these invariants, we codify our intuition by showing that
Theorem~\fullref{storeAtomicity} follows naturally from these
invariants.  A store may happen only when leaf cache is in \Mo{} state
for that address (Invariant~\fullref{processReq}). Leaf caches cannot
be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
other leaf caches will be in \In{} state preventing the possibility of
load and store requests for the same address (by Invariant
\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.

Given Invariant~\fullref{noReqAgain}, once a response has been produced
for a request $i$, the cache may subsequently only process $j^{th}$
request such that $j > i$. Thus there is at most one response to each
request, satisfying Definition~\fullref{uniqRespLabels}. A load request
may be processed only if a cache is at least in \Sh{} state (Invariant
\fullref{processReq}). As leaf caches have no children, a leaf cache in
at least in \Sh{} state for address $a$ is \clean for that address. A
load response supplies the data for that address present in the cache,
which, by Invariant~\fullref{lastestValue}, contains the ``latest
value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
is also satisfied. All that remains is to show that these properties hold.

@@

We can directly show that Invariants~\fullref{processReq} and
\fullref{noReqAgain} hold at the initial configuration and are
preserved through each atomic transition of our protocol
(Figure~\ref{trans}). Invariants~\fullref{uniqM} and
\fullref{latestValue} correctness rely on properties of the node
hierarchy. We can prove both using the following invariants on
parent-child relationships.

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \fullref{conservative} and
\fullref{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \fullref{conservative} and the first part
\fullref{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

@@

To prove Invariant \fullref{conservative} at some time $t$, consider the case
when a child cache $c$ has sent to or received from the parent $p$ some message
before $t$. Consider the last such message $m$ sent by or received at $c$ at
$t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
from $c$ after sending $m$ in order to not violate Invariant \fullref{noCross}.
Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
sending $m$ in order to not violate Invariants \fullref{noCross} and
\fullref{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
the response messages sync up the next-state according to the transitions in
Figure \ref{trans}). Since $p$ can not receive any message between $t''+1$ and
$t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
between $t'+1$ and $t$, Invariant \fullref{conservative} holds at $t$. If no
message has been sent or received by $c$ before $t$, then $p$
wouldn't have received any message sent by $c$ before $t$ as it means that the
message must be sent by $c$ before $t$. Since the initial states of the
directory and the child's state are both set to \In{}, the same argument holds
(in which the parent can only upgrade), thus maintaining Invariant
\fullref{conservative}.

@@

Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
other response can be received by the parent at time $t_p$ such that $t+1 \le
t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
directory for $c$. Similarly, no response can be sent by the child at time
$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
child upgrades its state at $t'$. Using arguments similar to the proof of
Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
state for address $a$ when a response is in transit from $p$ to $c$.
Similarly, we can prove that no cache can be in \Mo{} state for address $a$
when a response is in transit for address $a$ from $c$ to $p$. Since data is in
transit at the same time as when the response is in transit, Invariant
\fullref{noTransitWrite} follows.

@@

%Invariant~\fullref{conservative} can be proved by

@@

%Invariant~\fullref{conservative} at any time $t$ for child cache $c$ and parent
